A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-13 paper.tex
minor edits

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential thermodynamical interpretations of 
dust-gas drag dynamics, for example the streaming instability 
as a manifestation of pulsational instabilities in stars. 
\end{abstract}

@@

By re-casting the dusty problem into identical form to standard
hydrodynamics, we can follow standard procedures to generalize previous stability 
results for pure gas disks to the dusty case. We demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We
show that sharp dust edges may destabilize protoplanetary disks.  
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}

@@

By re-casting the dusty problem into identical form to 
hydrodynamics, we can apply standard methods to generalize previous stability 
results for pure gas disks to the dusty case. We demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We
show that sharp dust edges may destabilize protoplanetary disks.  
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations.
